Mapping malaria in pre-elimination settings by sharing spatial information between incidence and prevalence datasets

### 2019-08-23 ms.tex
Susan edits.
Amelia edits.

\title[Mapping malaria by sharing spatial information]{Mapping malaria in pre-elimination settings by sharing spatial information between incidence and prevalence datasets}
\author[Tim C.D. Lucas {\it et al.}]{Tim C.D. Lucas}
\address{Big Data Institute,
University of Oxford,
Oxford,
UK.}
\email{timcdlucas@gmail.com}
\author{Anita K. Nandi\textsuperscript{1}}
\author{Elisabeth G. Chestnutt\textsuperscript{1}}
\author{Katherine A. Twohig\textsuperscript{1}}
\author{Suzanne H. Keddie\textsuperscript{1}}
\author{Emma L. Collins\textsuperscript{1}}
\author{Rosalind E. Howes}
\author{Michele Nguyen\textsuperscript{1}}
\author{Susan Rumisha\textsuperscript{1}}
\author{Andre Python\textsuperscript{1}}
\author{Rohan Arambepola\textsuperscript{1}}
\author{Amelia Bertozzi-Villa\textsuperscript{1}}
\author{Penelope Hancock\textsuperscript{1}}
\author{Punam Amratia\textsuperscript{1}}
\author{Katherine E. Battle\textsuperscript{1}}
\author{Ewan Cameron\textsuperscript{1}}
\author{Pete W. Gething\textsuperscript{1, 2, 3}}
\author{Daniel J. Weiss\textsuperscript{1}}
\address{1. Big Data Institute,
University of Oxford,
Oxford,
UK.}
\address{2. Telethon Kids Institute, Perth Children’s Hospital, Perth, Australia}
\address{3. Curtin University, Perth, Australia}

@@

% Please keep the abstract below 300 words
\begin{abstract}
As malaria incidence decreases and more countries move towards elimination, maps of malaria risk in low prevalence areas are becoming increasingly needed.
For low burden areas, disaggregation regression models have been developed to estimate risk at high spatial resolution from routine surveillance reports of incidence aggregated to administrative unit polygons.
However, in areas with both routine surveillance data and prevalence surveys, models that make use of the spatial information from prevalence point surveys have great potential.
Using case studies in Indonesia, Senegal and Madagascar, we compare two methods for incorporating point-level, spatial information into disaggregation regression models.
The first simply fits a Gaussian random field to prevalence point-surveys to create a new covariate.
The second is a multi-likelihood model that jointly fits a model using prevalence point-surveys and polygon incidence data.
We find that the simple model generally performs better than a baseline disaggregation model while the joint model performance was mixed.
More generally, our results demonstrate that combining these types of data improves estimates of malaria incidence.
\end{abstract}

@@

Global malaria incidence has decreased dramatically over the last 20 years \citep{bhatt2015effect, weiss2019mapping, battle2019mapping}.
This decrease has been accompanied by a strategic shift aiming for elimination in low incidence countries \citep{world2016world, newby2016path}.
Accurate, high-resolution maps of malaria risk are vital in countries in the elimination and pre-elimination phases as they highlight areas with ongoing \emph{Plasmodium} transmission most in need of interventions \citep{sturrock2016mapping, cohen2017mapping}.
Mapping malaria in low burden countries presents new challenges as traditional mapping of prevalence \citep{gething2011new, bhatt2017improved, gething2012long, bhatt2015effect} using cluster-level surveys and model-based geostatistics are not necessarily effective in these areas \citep{sturrock2016mapping, sturrock2014fine}.
In low burden areas, very large sample sizes are needed before a prevalence survey is informative because so few individuals have parasitaemia that most sample points will have no cases.
However, the availability and quality of routine surveillance data of malaria case counts, typically aggregated administrative unit polygons, is improving, thus providing an alternative data source for mapping malaria burden \citep{sturrock2016mapping, ohrt2015information, cibulskis2011worldwide}.
Advantageously, the routine surveillance data can be more sensitive than prevalence point-surveys in low transmission areas because the entire public health system is being used to passively monitor disease risk \citep{cibulskis2011worldwide}.

@@

We used two data sources that reflect malaria burden; prevalence point-surveys and polygon incidence data.
Prevalence point surveys consist of geo-located household survey clusters wherein all sampled individuals are tested for malaria and the positive cases as well as the total number of children tested is recorded. 
Polygon incidence data is aggregated by administrative-level, typically by governmental organizations tasked with collating and summarizing data reported from hospitals and health facilities. 
Unlike the point data, polygon-level incidence only includes the number of cases and not the number of people relying on that unit’s healthcare facilities.  
As such, to determine an incidence rate we rely on gridded population surfaces, summarised using the administrative unit’s boundaries, to provide the denominator. 
The prevalence point-survey data were extracted from the Malaria Atlas Project prevalence survey database  \citep{bhatt2015effect, guerra2007assembling, pfeffer2018ma}. 
As the prevalence point-surveys cover different age ranges they were standardised to an age-range of 2--10 using the model from \citep{smith2007standardizing}. 
The age standardisation model gives the surveys with zero positive cases a small positive prevalence. 
The polygon incidence data were collated from various government reports and adjusted to a standard metric of malaria incidence using methods defined in \citep{cibulskis2011worldwide, weiss2019mapping}. 
These adjustments account for underreporting of clinical cases due to lack of treatment seeking, missed case reports, and cases that sought medical attention outside the public health systems \citep{battle2016treatment}. 
Where species specific reports were given, these were used, and in reports that did not distinguish between species of \emph{Plasmodium} the national estimate of the ratio between \emph{P. falciparum} and \emph{P. vivax} cases were used to estimate \emph{P. falciparum} only cases. 
These adjustments are uniform across each country. 
The polygon incidence data can be seen in Figures~\ref{predobsmapsen}--\ref{predobsmapidn}.

We selected Indonesia, Senegal and Madagascar as case examples as they all have abundant subnational surveillance data and country wide surveys from approximately the same period.
To minimise temporal effects we selected one year of polygon incidence data and the surrounding five years of prevalence point-survey data for each country.
For Indonesia we selected polygon incidence data from 2012 that covers 379 regions, and prevalence data from 2010 to 2014 that consists of 1,233 surveys, representing 230,747 individuals.
For Senegal we selected 2015 for polygon incidence data (41 regions) and 2013 to 2017 for prevalence data (804 surveys, 17,037 individuals).
Finally, for Madagascar we selected 2013 for polygon incidence (110 regions) and 2011 to 2015 for prevalence data (1,049 surveys, 36,411 individuals).

@@

where $i_a$ is the estimated polygon incidence rate and $\mathrm{pop_a}$ is the (assumed to be) known polygon population-at-risk.

@@

>>>>

\title[Mapping malaria by sharing spatial information]{Mapping malaria in pre-elimination settings by sharing spatial information between incidence and prevalence datasets}
\author[Tim C.D. Lucas {\it et al.}]{Tim C.D. Lucas}
\address{Big Data Institute,
University of Oxford,
Oxford,
UK.}
\email{timcdlucas@gmail.com}
\author{Anita K. Nandi\textsuperscript{1}}
\author{Elisabeth G. Chestnutt\textsuperscript{1}}
\author{Katherine A. Twohig\textsuperscript{1}}
\author{Suzanne H. Keddie\textsuperscript{1}}
\author{Emma L. Collins\textsuperscript{1}}
\author{Rosalind E. Howes}
\author{Michele Nguyen\textsuperscript{1}}
\author{Susan F. Rumisha\textsuperscript{1}}
\author{Andre Python\textsuperscript{1}}
\author{Rohan Arambepola\textsuperscript{1}}
\author{Amelia Bertozzi-Villa\textsuperscript{1}}
\author{Penelope Hancock\textsuperscript{1}}
\author{Punam Amratia\textsuperscript{1}}
\author{Katherine E. Battle\textsuperscript{1}}
\author{Ewan Cameron\textsuperscript{1}}
\author{Peter W. Gething\textsuperscript{1, 2, 3}}
\author{Daniel J. Weiss\textsuperscript{1}}
\address{1. Big Data Institute,
University of Oxford,
Oxford,
UK.}
\address{2. Telethon Kids Institute, Perth Children’s Hospital, Perth, Australia}
\address{3. Curtin University, Perth, Australia}

@@

% Please keep the abstract below 300 words
\begin{abstract}
As malaria incidence decreases and more countries move towards elimination, maps of malaria risk in low prevalence areas are becoming increasingly needed.
For low burden areas, disaggregation regression models have been developed to estimate risk at high spatial resolution from routine surveillance reports of incidence aggregated to administrative unit polygons.
However, in areas with both routine surveillance data and prevalence surveys, models that make use of the spatial information from prevalence point surveys have great potential.
Using case studies in Indonesia, Senegal and Madagascar, we compare two methods for incorporating point-level, spatial information into disaggregation regression models.
The first simply fits a Gaussian random field to prevalence point-surveys to create a new covariate.
The second is a multi-likelihood model that is fitted jointly to prevalence point-surveys and polygon incidence data.
We find that the simple model generally performs better than a baseline disaggregation model while the joint model performance was mixed.
More generally, our results demonstrate that combining these types of data improves estimates of malaria incidence.
\end{abstract}

@@

Global malaria incidence has decreased dramatically over the last 20 years \citep{bhatt2015effect, weiss2019mapping, battle2019mapping}.
This decrease has been accompanied by a strategic shift aiming for elimination in low incidence countries \citep{world2016world, newby2016path}.
Accurate, high-resolution maps of malaria risk are vital in countries in the elimination and pre-elimination phases as they highlight areas with ongoing \emph{Plasmodium} transmission most in need of interventions \citep{sturrock2016mapping, cohen2017mapping}.
Mapping malaria in low burden countries presents new challenges as traditional mapping of prevalence \citep{gething2011new, bhatt2017improved, gething2012long, bhatt2015effect} using cluster-level surveys and model-based geostatistics are not necessarily effective in these areas \citep{sturrock2016mapping, sturrock2014fine}.
In low burden areas, very large sample sizes are needed before a prevalence survey is informative because so few individuals have parasitaemia that most sample points will have no cases.
In most cases, these large sample sizes are neither logistically nor financially feasible.
However, the availability and quality of routine surveillance data of malaria case counts, typically aggregated administrative unit polygons, is improving, thus providing an alternative data source for mapping malaria burden \citep{sturrock2016mapping, ohrt2015information, cibulskis2011worldwide}.
Advantageously, the routine surveillance data can be more sensitive than prevalence point-surveys in low transmission areas because the entire public health system is being used to passively monitor disease risk \citep{cibulskis2011worldwide}.

@@

We used two data sources that reflect malaria burden: prevalence point-surveys and polygon incidence data.
Prevalence point surveys consist of geo-located household survey clusters wherein all sampled individuals are tested for malaria and the positive cases as well as the total number of children tested is recorded. 
Polygon incidence data is aggregated by administrative-level, typically by governmental organizations tasked with collating and summarizing data reported from hospitals and health facilities. 
Unlike the point data, polygon-level incidence only includes the number of cases and not the number of people relying on that unit’s healthcare facilities.  
As such, to determine an incidence rate we rely on gridded population surfaces, summarised using the administrative unit’s boundaries, to provide the denominator. 
The prevalence point-survey data were extracted from the Malaria Atlas Project prevalence survey database  \citep{bhatt2015effect, guerra2007assembling, pfeffer2018ma}. 
As the prevalence point-surveys cover different age ranges they were standardised to an age-range of 2--10 using the model from \citep{smith2007standardizing}. 
The age standardisation model gives the surveys with zero positive cases a small positive prevalence. 
The polygon incidence data were collated from various government reports and adjusted to a standard metric of malaria incidence using methods defined in \citep{cibulskis2011worldwide, weiss2019mapping}. 
These adjustments account for underreporting of clinical cases due to lack of treatment seeking, missed case reports, and cases that sought medical attention outside the public health systems \citep{battle2016treatment}. 
Where species specific reports were given, these were used, and in reports that did not distinguish between species of \emph{Plasmodium} the national estimate of the ratio between \emph{P. falciparum} and \emph{P. vivax} cases were used to estimate \emph{P. falciparum} only cases. 
These adjustments are uniform across each country. 
The polygon incidence data can be seen in Figures~\ref{predobsmapsen}--\ref{predobsmapidn}.

We selected Indonesia, Senegal and Madagascar as case examples as they all have abundant subnational surveillance data and country wide surveys from approximately the same period.
To minimise temporal effects we selected one year of polygon incidence data and the surrounding five years of prevalence point-survey data for each country.
For Indonesia we selected polygon incidence data from 2012 that covers 379 regions, and prevalence data from 2010 to 2014 that consists of 1,233 survey clusters (i.e. unique locations), representing 230,747 individuals.
For Senegal we selected 2015 for polygon incidence data (41 regions) and 2013 to 2017 for prevalence data (804 surveys, 17,037 individuals).
Finally, for Madagascar we selected 2013 for polygon incidence (110 regions) and 2011 to 2015 for prevalence data (1,049 surveys, 36,411 individuals).

@@

where $i_a$ is the estimated polygon incidence rate and $\mathrm{pop_a}$ is the population-at-risk within that admin unit polygon (as apposed to the true health centre catchment area).

@@

\end{figure}
% todo add country names to plots.
